---
title: "GSEA"
format: html
editor: visual
---

## Tcga brca gsea hallmark

`GSEA on TCGA-BRCA ranked gene list (cancer vs. normal)`

`Focus: metabolic Hallmark gene sets`

```{r}
BiocManager::install("clusterProfiler")
```

```{r}
install.packages("msigdbr")
```

```{r}
library(clusterProfiler)
library(msigdbr)
library(dplyr)
library(ggplot2)
```

```{r}
#load data from python
ranked <- read.csv("ranked_gene_list_for_gsea.csv")
```

```{r}
# clusterProfiler wants a NAMED, SORTED numeric vector: names = gene symbols,
# values = your ranking statistic (DESeq2's Wald stat)
gene_list <- ranked$stat
names(gene_list) <- ranked$gene_symbol
gene_list <- sort(gene_list, decreasing = TRUE) #must be sorted descending

head(gene_list)
tail(gene_list)
```

### Get Hallmark gene sets

```{r}
hallmark_sets <- msigdbr(species = "Homo sapiens",category = "H") %>% dplyr::select(gs_name, gene_symbol)

# Quick look at which metabolic-relevant Hallmark sets exist
unique(hallmark_sets$gs_name)[grepl("GLYCOL|OXIDATIVE|FATTY|METABOL", unique(hallmark_sets$gs_name))]
```

```{r}
#run GSEA
gsea_res <- GSEA(
  geneList = gene_list,
  TERM2GENE = hallmark_sets,
  pvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500,
  seed = TRUE
)
```

Look at the full results table, sorted by significance

```{r}
gsea_df <- as.data.frame(gsea_res) %>% arrange(p.adjust)
print(gsea_df[, c("ID", "NES", "pvalue", "p.adjust")])
```

`NES = Normalized Enrichment Score`

`Positive NES  -> pathway genes skew toward the TOP of your ranked list (UP in cancer, since you ranked cancer-vs-normal with positive stat = higher in cancer)`

`Negative NES  -> pathway genes skew toward the BOTTOM (DOWN in cancer)`

`p.adjust  -> same idea as padj before`

We can check specific pathways

```{r}
targets <- c("HALLMARK_GLYCOLYSIS",
             "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
             "HALLMARK_FATTY_ACID_METABOLISM")

gsea_df %>% filter(ID %in% targets) %>%
  dplyr::select(ID, NES, pvalue, p.adjust)
```

Plot the classic GSEA enrichment curve for each

```{r}
library(enrichplot)
for (term in targets) {
  if (term %in% gsea_res@result$ID) {
    p <- gseaplot2(gsea_res, geneSetID = term, title = term)
    ggsave(paste0("gsea_plot_", term, ".png"), p, width = 8, height = 6, dpi = 150)
  }
}
```

```{r}
p_dot <- dotplot(gsea_res, showCategory = 20, split = ".sign") + facet_grid(. ~ .sign)
ggsave("gsea_dotplot_top20.png", p_dot, width = 10, height = 8, dpi = 150)

cat("\nDone. Check for:\n")
cat("  - gsea_results_hallmark.csv (full table)\n")
cat("  - gsea_plot_HALLMARK_*.png (individual enrichment curves)\n")
cat("  - gsea_dotplot_top20.png (summary figure)\n")
```
