---
title: "Subtype stratification"
format: html
---

## Subtype Stratification of Mitochondrial Pathway Scores

Tests whether the mitochondrial/metabolic shifts found in the whole-cohort comparison (Fatty Acid Oxidation down, biogenesis/import up) differ by molecular subtype (Luminal A, Luminal B, HER2-enriched, Basal-like).

Expectation from the literature: Basal-like/TNBC tumors typically show the strongest glycolytic shift and metabolic rewiring; Luminal tumors tend to show milder changes.

## Prepare the data with subtypes

```{r}
library(TCGAbiolinks)
library(dplyr)
```

```{r}
subtype_info <- TCGAquery_subtype(tumor = "BRCA")

# Look at what's actually in it
colnames(subtype_info)
head(subtype_info)
```

```{r}
table(subtype_info$BRCA_Subtype_PAM50, useNA = "always")
```

```{r}
sample_metadata <- read.csv("sample.csv")
```

```{r}
# Check actual structure
str(subtype_info)
colnames(subtype_info)
head(subtype_info$patient, 10)  # or whatever the actual barcode column is called
```

```{r}
head(subtype_info$patient, 10)
```

```{r}
# patient matches with tcga.gdc_cases.submitter_id in sample_metadata
head(sample_metadata$tcga.gdc_cases.submitter_id, 10)
```

```{r}
merged_test <- sample_metadata %>%
  left_join(
    subtype_info %>% dplyr::select(patient, BRCA_Subtype_PAM50),
    by = c("tcga.gdc_cases.submitter_id" = "patient")
  )

#Check 
"BRCA_Subtype_PAM50" %in% colnames(merged_test)
dim(merged_test)   # should have same row count as sample_metadata, one extra column
table(merged_test$BRCA_Subtype_PAM50, useNA = "always")
```

```{r}
sample_metadata <- merged_test
```

```{r}
#some normal tissus were labled because were obtained from same cancer patient. here we will label normall tissues with NA
sample_metadata <- sample_metadata %>%
  mutate(BRCA_Subtype_PAM50 = ifelse(
    tcga.cgc_sample_sample_type == "Solid Tissue Normal",
    NA,
    BRCA_Subtype_PAM50
  ))

# Recheck
table(sample_metadata$BRCA_Subtype_PAM50, useNA = "always")
```

```{r}
#we will remove tumor sample with na subtype for analysis 
subtype_analysis_data <- sample_metadata %>%
  filter(tcga.cgc_sample_sample_type == "Primary Tunor",
         !is.na(BRCA_Subtype_PAM50))
```

```{r}
cat("Tumor samples with valid subtype:", nrow(subtype_analysis_data), "\n")
cat("Tumor samples excluded (missing subtype):",
    sum(sample_metadata$tcga.cgc_sample_sample_type == "Primary Tumor" &
          is.na(sample_metadata$BRCA_Subtype_PAM50)), "\n")
```

```{r}
library(dplyr)

# Check how many rows match the condition you want to remove, before removing
sample_metadata %>%
  filter(tcga.cgc_sample_sample_type == "Primary Tumor", is.na(BRCA_Subtype_PAM50)) %>%
  nrow()

# Remove those specific rows: keep everything EXCEPT
# (Primary Tumor AND missing subtype)
sample_metadata <- sample_metadata %>%
  filter(!(tcga.cgc_sample_sample_type == "Primary Tumor" & is.na(BRCA_Subtype_PAM50)))

# Confirm the result
nrow(sample_metadata)
table(sample_metadata$tcga.cgc_sample_sample_type, useNA = "always")
table(sample_metadata$BRCA_Subtype_PAM50, useNA = "always")
```

```{r}
ncol(sample_metadata)
nrow(sample_metadata)
```

```{r}
write.csv(sample_metadata,"sample_metadata_with_subtype.csv", row.names=TRUE)
```

```{r}
data <- read.csv("sample_metadata_with_subtype.csv")
ncol(data)
nrow(data)
```

## Subtype Stratification 

```{python}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy.stats import kruskal
import scikit_posthocs as sp
import seaborn as sns
```

```{python}
sample_meta= pd.read_csv("sample_metadata_with_subtype.csv")
pathway_scores = pd.read_csv("mitocarta_pathway_subscores.csv", index_col=0)
mito_score = pd.read_csv("mitochondrial_score_per_sample.csv", index_col=0).squeeze()
```

```{python}
subtype_col = "BRCA_Subtype_PAM50"
 
print("Subtype counts (all samples, including NA normals):")
print(sample_meta[subtype_col].value_counts(dropna=False))
```

```{python}
tumor_with_subtype = sample_meta[sample_meta[subtype_col].notna()].copy()
print(f"\nTumor samples with a valid PAM50 subtype: {len(tumor_with_subtype)}")
print(tumor_with_subtype[subtype_col].value_counts())
```

```{python}
tumor_with_subtype = tumor_with_subtype.set_index("tcga.tcga_barcode")
```

```{python}
print(tumor_with_subtype.index[:5])
print(mito_score.index[:5])
print(pathway_scores.index[:5])
```

```{python}
#Attach scores
tumor_with_subtype["mito_score"] = mito_score.reindex(tumor_with_subtype.index)
tumor_with_subtype = tumor_with_subtype.join(
    pathway_scores.reindex(tumor_with_subtype.index)
)
```

```{python}
n_matched = tumor_with_subtype["mito_score"].notna().sum()
print(f"Samples with a matched mito_score after join: {n_matched} out of {len(tumor_with_subtype)}")
assert n_matched > len(tumor_with_subtype) * 0.9, (
    "Fewer than 90% of samples matched -- check index formats between "
    "tumor_with_subtype, mito_score, and pathway_scores before proceeding."
)
 
print("\nPer-subtype sample count after join (should match the raw value_counts):")
print(tumor_with_subtype.groupby(subtype_col)["mito_score"].apply(lambda x: x.notna().sum()))
```

## Kruskal-Wallis

Kruskal-Wallis = non-parametric equivalent of one-way ANOVA, appropriate here since we shouldn't assume the scores are normally distributed within each subtype group, and group sizes are unequal.

```{python}
subtype_groups = [
    tumor_with_subtype.loc[tumor_with_subtype[subtype_col] == s, "mito_score"].dropna()
    for s in tumor_with_subtype[subtype_col].unique()
]
stat, pval = kruskal(*subtype_groups)
print(f"\nKruskal-Wallis test, overall mitochondrial score across subtypes: "
      f"H={stat:.2f}, p={pval:.2e}")
```

means:

- **Null hypothesis (H₀):** mitochondrial scores are the same across all breast cancer subtypes

- **Alternative hypothesis (H₁):** at least one subtype has a different mitochondrial score distribution

  There is a statistically significant difference in overall MitoCarta mitochondrial pathway scores among breast cancer subtypes.

## Dunn's test

```{python}
# since Kruskal-Wallis only tells you "groups differ somewhere", not which
dunn_result = sp.posthoc_dunn(
  tumor_with_subtype, val_col="mito_score", group_col=subtype_col, p_adjust="fdr_bh"
)
print("\nPairwise subtype comparisons (Dunn's test, FDR-adjusted p-values):")
print(dunn_result)
```

- **LumB is different from every other group** (all FDR \< 0.001)

- **Normal-like is different from every subtype** (all FDR \< 0.05)

- **Basal and HER2 are not distinguishable from each other**

- **Basal and HER2 are not significantly different from LumA** (although LumA vs Basal is trending)

  LumB tumors show a distinct mitochondrial transcriptional program compared with other breast cancer subtypes.

```{python}
print(tumor_with_subtype.groupby(subtype_col)["mito_score"].agg(["mean", "median", "count"]).sort_values("mean"))
```

```{python}
#Visualize overall score by subtype
plt.figure(figsize=(9, 6))
order = tumor_with_subtype.groupby(subtype_col)["mito_score"].median().sort_values().index
sns.boxplot(data=tumor_with_subtype, x=subtype_col, y="mito_score", order=order)
sns.stripplot(data=tumor_with_subtype, x=subtype_col, y="mito_score", order=order,
              color="black", alpha=0.3, size=2)
plt.title("Overall mitochondrial score by PAM50 subtype")
plt.tight_layout()
#plt.savefig("mito_score_by_subtype.png", dpi=150)
plt.show()
```

```{python}
#test specific pathway
target_pathway = "Fatty acid oxidation"
 
if target_pathway in tumor_with_subtype.columns:
    fao_groups = [
        tumor_with_subtype.loc[tumor_with_subtype[subtype_col] == s, target_pathway].dropna()
        for s in tumor_with_subtype[subtype_col].unique()
    ]
    stat_fao, pval_fao = kruskal(*fao_groups)
    print(f"\nKruskal-Wallis, {target_pathway} across subtypes: "
          f"H={stat_fao:.2f}, p={pval_fao:.2e}")
 
    print(f"\nMean {target_pathway} score by subtype:")
    print(tumor_with_subtype.groupby(subtype_col)[target_pathway].mean().sort_values())
```

- **Overall mitochondrial gene expression** (biogenesis-heavy, all \~1,058 genes combined) does not cleanly separate Basal from other subtypes. LumB stands out instead, likely tracking proliferation rate rather than oxidative capacity specifically.

- **Fatty acid oxidation specifically** a narrower, functionally coherent pathway does show the clean, strong, subtype-specific pattern predicted by the literature, with Basal-like tumors showing the sharpest suppression.

mitochondrial dysfunction in cancer story isn't one uniform phenomenon, it depends heavily on which mitochondrial function you're measuring. Broad biogenesis measures reflect proliferation differences across subtypes; specific oxidative fuel-use pathways like fatty acid oxidation reflect the actual metabolic rewiring pattern tied to subtype biology (Basal/TNBC tumors are well known in the literature for relying more heavily on glycolysis and having reduced fatty acid oxidative capacity)

## Lumb Pathway drivers

```{python}
from scipy.stats import mannwhitneyu
from statsmodels.stats.multitest import multipletests

subtype_col = "BRCA_Subtype_PAM50"
pathway_cols = [c for c in pathway_scores.columns]   # from your existing script
```

```{python}
#For each pathway, compare LumB vs. all other subtypes combined
is_lumb = tumor_with_subtype[subtype_col] == "LumB"
 
results = []
for pathway in pathway_cols:
    lumb_vals = tumor_with_subtype.loc[is_lumb, pathway].dropna()
    other_vals = tumor_with_subtype.loc[~is_lumb, pathway].dropna()
 
    if len(lumb_vals) < 5 or len(other_vals) < 5:
        continue
 
    stat, pval = mannwhitneyu(lumb_vals, other_vals)
    effect_size = lumb_vals.mean() - other_vals.mean()   # positive = higher in LumB
 
    results.append({
        "pathway": pathway,
        "lumb_mean": lumb_vals.mean(),
        "other_mean": other_vals.mean(),
        "difference": effect_size,
        "pvalue": pval,
    })
 
results_df = pd.DataFrame(results)
results_df["padj"] = multipletests(results_df["pvalue"], method="fdr_bh")[1]
results_df = results_df.sort_values("difference", ascending=False)
 
results_df.to_csv("lumb_vs_other_pathway_drivers.csv", index=False)
```

```{python}
#Show the pathways most ELEVATED in LumB (driving it UP)
print("=== Top 15 pathways MOST ELEVATED in LumB (driving the high score) ===")
top_up = results_df[results_df["padj"] < 0.05].head(15)
print(top_up[["pathway", "lumb_mean", "other_mean", "difference", "padj"]].to_string())
```

```{python}
#Also show the pathways most SUPPRESSED in LumB
print("\n=== Top 15 pathways MOST SUPPRESSED in LumB ===")
bottom = results_df[results_df["padj"] < 0.05].tail(15).sort_values("difference")
print(bottom[["pathway", "lumb_mean", "other_mean", "difference", "padj"]].to_string())
```

Luminal B breast tumors exhibit coordinated upregulation of mitochondrial protein import, translation, and cristae organization, supporting enhanced oxidative phosphorylation and fatty acid oxidation, while simultaneously downregulating antioxidant and iron homeostasis pathways, suggesting a high-energy but potentially redox-vulnerable mitochondrial state.
