---
title: "08_wgcna"
format: html
---

## WGCNA: Weighted Gene Co-expression Network Analysis

Instead of testing each gene individually against a trait (like DESeq2 step), WGCNA finds groups of genes ("modules") whose expression rises and falls TOGETHER across samples, genes acting as a coordinated unit, which often reflects shared regulation or shared biological function. Each module then gets correlated with traits of interest (subtype, mitochondrial score, etc.) as a WHOLE, rather than gene-by-gene.

This is a genuinely different question than DESeq2/GSEA answered: not "which genes changed," but "which genes move together, and does that coordinated group relate to subtype or mitochondrial function.

```{r}
install.packages("WGCNA")
```

```{r}
library(WGCNA)
library(dplyr)

options(stringsAsFactors = FALSE)
allowWGCNAThreads()
```

load the expression data

WGCNA wants samples as rows, genes as columns, the opposite orientation from DESeq2. Also we will use variance-stabilized or log-normalized expression, NOT raw counts.

```{r}
counts_filtered <- read.csv("data_filtered.csv", row.names = 1)
gene_metadata <- read.csv("tcga_brca_gene_metadata.csv", row.names =1)
```

```{r}
log_expr <- log2(counts_filtered + 1)
expr_for_wgcna <- as.data.frame(t(log_expr))
```

Filter to the most informative genes before running wgcna

WGCNA on \~15,000+ genes is computationally heavy and mostly adds noise, standard practice is to restrict to the most variable genes (the ones actually carrying information about differences between samples). Common choice: top 5,000-ish most variable genes.

```{r}
gene_var <- apply(expr_for_wgcna, 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
expr_for_wgcna <- expr_for_wgcna[, top_genes]
```

Check for outlier samples BEFORE building the network

```{r}
pca <- prcomp(expr_for_wgcna, scale. = TRUE)
pca_df <- as.data.frame(pca$x[, 1:2])

# Flag samples more than ~3-4 standard deviations from the mean on PC1 or PC2
pc1_outliers <- rownames(pca_df)[abs(scale(pca_df$PC1)) > 4]
pc2_outliers <- rownames(pca_df)[abs(scale(pca_df$PC2)) > 4]
print(c(pc1_outliers, pc2_outliers))
```

no samples crossed that extreme outlier threshold.

Choose a "soft threshold" power

WGCNA needs one tuning parameter (power) that determines how strongly to emphasize strong correlations over weak ones when building the network. This step tests a range of values and picks the smallest one that achieves approximate "scale-free" network structure a property real biological networks tend to have.

```{r}
powers <- c(1:20)
sft <- pickSoftThreshold(expr_for_wgcna, powerVector = powers,
                         verbose = 5, networkType = "signed")

pdf("wgcna_soft_threshold.pdf", width = 10, height = 5)
par(mfrow = c(1, 2))
plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)", ylab = "Scale Free Topology R^2",
     main = "Scale independence")
abline(h = 0.85, col = "red")
plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
     main = "Mean connectivity")
dev.off()
```

```{r}
print(sft$fitIndices[, c(1,2,5)])
```

R² crosses above 0.85 at **power = 13** (R²=0.90), power 12 is still just under (0.838). Standard WGCNA practice is to pick the smallest power that clears the threshold, so **13**

Build the network and detect modules

```{r}
chosen_power <- 13

net <- blockwiseModules(
  expr_for_wgcna,
  power = chosen_power,
  networkType = "signed",
  TOMType = "signed",
  minModuleSize = 30,
  reassignThreshold = 0,
  mergeCutHeight = 0.25,
  numericLabels = TRUE,
  saveTOMs = FALSE,
  verbose = 3
)

module_colors <- labels2colors(net$colors)
table(module_colors)
```

we have 9 modules total, grey isn't a real module, it's genes that didn't cluster into any coherent group.

\
Visualize the module dendrogram

```{r}
plotDendroAndColors(net$dendrograms[[1]], module_colors[net$blockGenes[[1]]],
                    "Module colors", dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```

do these genes overlap with known pathways more than chance?

To answer "which genes, and do they share function". We will extract each module's gene list, then run pathway enrichment **per module**

```{r}
module_assignments <-data.frame(
  gene = colnames(expr_for_wgcna),
  module = module_colors
)

write.csv(module_assignments, "wgcna_module_assignments.csv", row.names = FALSE)
```
