---
title: "10_Wgcna_trait_correlation"
format: html
---

## Wgcna module trait correlation

A module eigengene (ME) is a single summary value per sample per module, essentially "the first principal component" of that \# module's genes. Instead of 1,274 genes in turquoise, you get ONE number per sample representing that module's overall activity.

This lets you correlate whole modules against each other and against traits directly, answering:

\- Does the proliferation module (green) correlate with metabolic scores, supporting "proliferation drives metabolic rewiring"?

\- Does the immune module (yellow) correlate with metabolic scores, supporting "immune cell metabolism contributes to the signal"?

\- Does the hormone module (brown) correlate with metabolic scores?

```{r}
library(WGCNA)
library(dplyr)
```

## Step 1: Recompute module eigengenes from your existing network

```{r}
counts_filtered <- read.csv("data_filtered.csv", row.names = 1)
log_expr <- log2(counts_filtered + 1)
expr_for_wgcna_full <- as.data.frame(t(log_expr))
```

```{r}
# Re-apply the SAME top-5000-most-variable-genes filter used in Step 1
gene_var <- apply(expr_for_wgcna_full, 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
expr_for_wgcna <- expr_for_wgcna_full[, top_genes]

# Now this should have exactly 5000 columns, matching module_colors
cat("expr_for_wgcna columns:", ncol(expr_for_wgcna), "\n")
cat("module_colors length:", length(module_colors), "\n")
# These two numbers MUST match before proceeding
```

```{r}
MEs <- moduleEigengenes(expr_for_wgcna, colors =module_colors )$eigengenes
MEs <- orderMEs(MEs)# reorders so similar modules sit near each other
MEs <- MEs[,colnames(MEs) !="MEgrey"]
```

```{r}
head(MEs,5)
```

##Step 2: Module-vs-module correlation (does proliferation track with metabolism, does immune track with metabolism, etc.)

```{r}
me_cor <- cor(MEs, use = "pairwise.complete.obs")
me_cor_pval <- corPvalueStudent(me_cor, nSamples = nrow(MEs))
```

```{r}
head(me_cor_pval, 5)
```

```{r}
labeledHeatmap(
  Matrix = me_cor,
  xLabels = colnames(MEs),
  yLabels = colnames(MEs),
  colorLabels = FALSE,
  colors = blueWhiteRed(50),
  textMatrix = round(me_cor, 2),
  setStdMargins = FALSE,
  cex.text = 0.6,
  main = "Module-Module Eigengene Correlation"
)
```

## Step 3: Bring existing MitoCarta scores + clinical traits

```{r}
sample_meta <- read.csv("sample.csv", row.names = 1)
mito_score <- read.csv("mitochondrial_score_per_sample.csv", row.names = 1)
pathway_scores <- read.csv("mitocarta_pathway_subscores.csv", row.names = 1)
```

align everything to the same sample set as MEs

```{r}
rownames(MEs) <- gsub("\\.", "-", rownames(MEs))
common_samples <- intersect(rownames(MEs), rownames(sample_meta))
MEs_aligned <- MEs[common_samples,]
```

build trait matrix: tumor/normal (binary), subtype, FAO score, overall mito score

```{r}
traits <- data.frame(row.names = common_samples)
traits$is_tumor <- ifelse(sample_meta[common_samples, "tcga.cgc_sample_sample_type"] == "Primary Tumor", 1, 0)
traits$mito_score <- mito_score[common_samples,1]
```

```{r}
module_trait_cor <- cor(MEs_aligned, traits, use = "pairwise.complete.obs")
module_trait_pval <- corPvalueStudent(module_trait_cor, nSamples = nrow(traits))
```

```{r}
module_trait_cor
module_trait_pval 
```

As shown in the table, the green module (cell cycle/proliferation genes: E2F Targets, G2M Checkpoint) shows a moderately strong positive correlation with tumor status (r=0.60) and a weaker but still positive correlation with the overall mitochondrial score (r=0.29) — consistent with proliferation being coupled to increased mitochondrial biogenesis. The black module shows the strongest correlation with mitochondrial score of any module (r=0.55, moderate-to-moderately-strong), despite its earlier enrichment results (tight junction, apoptosis, estrogen response, with glycolysis as a minor secondary hit) not obviously pointing to a mitochondrial identity — this relationship is statistically robust (p\~1e-96) but unexplained by the module's current annotation, and worth investigating further (by examining its hub genes) rather than treated as resolved.
