---
title: "13_estimate"
format: html
editor: visual
---

Tumors with more immune cells might naturally have:

- Higher **ICR score** (more immune genes)

- And possibly different **FAO levels**

So the concern is: Maybe FAO and ICR look correlated just because some tumors have more immune cells overall.

What partial correlation does? It **removes the effect of immune cell abundance** (using ESTIMATE ImmuneScore)

**ESTIMATE ImmuneScore:** It is a transcriptomic estimate of how much immune-cell signal is present in a bulk tumor sample

## ESTIMATE Setup

```{r}
# ESTIMATE isn't on CRAN — install from R-Forge
install.packages("estimate", repos = "http://r-forge.r-project.org", type = "source")
```

```{r}
library(estimate)
```

```{r}
expr_matrix <- read.csv("counts_basal_normed.csv")
```

```{r}
head(expr_matrix, 5)
```

```{r}
# Move the barcode column into rownames
rownames(expr_matrix) <- expr_matrix$tcga.tcga_barcode

# Drop it so it's not treated as a "gene"
expr_matrix$tcga.tcga_barcode <- NULL

# NOW transpose — genes end up as rows, real barcodes as columns
expr_matrix_t <- t(expr_matrix)

#head(expr_matrix_t, 5)    # check what these look like now
```

```{r}
write.table(expr_matrix_t, "basal_expr_for_estimate.txt",
            sep = "\t", quote = FALSE, col.names = NA)

# filter to the ~10,000 common genes ESTIMATE uses, output a .gct file
filterCommonGenes(
  input.f = "basal_expr_for_estimate.txt",
  output.f = "basal_expr_filtered.gct",
  id = "GeneSymbol"
)

#compute scores
estimateScore(
  input.ds = "basal_expr_filtered.gct",
  output.ds = "basal_estimate_scores.gct",
  platform = "illumina"   # correct default for RNA-seq data, not just microarray
)
```

```{r}
scores_check <- read.table("basal_estimate_scores.gct", sep = "\t", skip = 2, header = TRUE, row.names = 1)
dim(scores_check)   # should be 4 rows (Stromal, Immune, ESTIMATE, TumorPurity) x n_samples
head(scores_check[, 1:5])
```
